<a id="top"></a>

[![Build Status](https://travis-ci.com/vrunge/gfpop.svg?branch=master)](https://travis-ci.com/vrunge/gfpop)
[![](https://img.shields.io/badge/docs-vignettes-blue.svg)](https://github.com/vrunge/gfpop)

<!-- 
%\VignetteEngine{knitr::rmarkdown} 
%\VignetteIndexEntry{An Introduction to gfpop}
--> 

# gfpop Vignette
### Vincent Runge
#### LaMME, Evry University
### March 2, 2022

> [Quick Start](#qs)

> [Some examples](#se)

> [Graph construction](#gc)

> [Supplementary R functions](#suppl)


## Quick Start

we present a basic use of the main functions of the `gfpop` package. More details about optional arguments are given in later sections.

We install the package from Github:

```{r}
#devtools::install_github("vrunge/gfpop")
library(gfpop)
```

We simulate some univariate gaussian data (`n = 1000` points) with relative change-point positions `0.1, 0.3, 0.5, 0.8, 1` and means `1, 2, 1, 3, 1` with a variance equal to `1`.

```{r}
n <- 1000
myData <- dataGenerator(n, c(0.1,0.3,0.5,0.8,1), c(1,2,1,3,1), sigma = 1)
```

We define the graph of constraints to use for the dynamic programming algorithm. A simple case is the up-down constraint with a penalty here equal to a classic `2 log(n)`.

```{r}
myGraph <- graph(penalty = 2*log(n), type = "updown")
```

The gfpop function gives the result of the segmentation using `myData` and `myGraph` as parameters. We choose a gaussian cost.

```{r}
gfpop(data = myData, mygraph = myGraph, type = "mean")
```

The vector `changepoints` gives the last index of each segment. It always ends with the length of the vector `vectData`.

The vector `states` contains the states in which lies each mean. The length of this vector is the same as the length of `changepoint`.

The vector `forced` is a boolean vector. A forced element means that two consecutive means have been forced to satisfy the constraint. For example, the "up" edge with parameter c is forced if m(i+1) - m(i) = c.

The vector `parameters` contains the inferred means/parameters of the successive segments. 
 
The number `globalCost` is equal to the non-penalized cost, that is the value of the fit to the data ignoring the penalties for adding changes.

<a id="se"></a>

## Some examples

### Isotonic regression

The isotonic regression infers a sequence of nondecreasing means. 


```{r}
n <- 1000
mydata <- dataGenerator(n, c(0.1, 0.2, 0.3, 0.4, 0.6, 0.8, 1), c(0, 0.5, 1, 1.5, 2, 2.5, 3), sigma = 1)
myGraphIso <- graph(penalty = 2*log(n), type = "isotonic")
gfpop(data =  mydata, mygraph = myGraphIso, type = "mean")
```

In this example, we use in `gfpop` function a robust biweight gaussian cost with `K = 1` and the `min` parameter in order to infer means greater than `0.5`.

### Fixed number of change-points

This algorithm is called segment neighborhood in the change-point litterature. In this example, we fixed the number of segments at 3 with an isotonic constraint. The graph contains two "up" edges with no cycling.


```{r}
n <- 1000
mydata <- dataGenerator(n, c(0.1, 0.2, 0.3, 0.4, 0.6, 0.8, 1), c(0, 0.5, 1, 1.5, 2, 2.5, 3), sigma = 1)
beta <- 0
myGraph <- graph(
  Edge(0, 1,"up", beta),
  Edge(1, 2, "up", beta),
  Edge(0, 0, "null"),
  Edge(1, 1, "null"),
  Edge(2, 2, "null"),
  StartEnd(start = 0, end = 2))

gfpop(data =  mydata, mygraph = myGraph, type = "mean")
```

### Robust up-down with constrained starting and ending states

In presence of outliers we need a robust loss (biweight). We can also force the starting and ending state and a minimal gap between the means (here equal to `1`)

```{r}
n <- 1000
chgtpt <- c(0.1, 0.3, 0.5, 0.8, 1)
myData <- dataGenerator(n, chgtpt, c(0, 1, 0, 1, 0), sigma = 1)
myData <- myData + 5 * rbinom(n, 1, 0.05) - 5 * rbinom(n, 1, 0.05)
beta <- 2 * log(n)
myGraph <- graph(
         Edge("Dw", "Up", type = "up", penalty = beta, gap = 1, K = 3),
         Edge("Up", "Dw", type = "down", penalty = beta, gap = 1, K = 3),
         Edge("Dw", "Dw", type = "null", K = 3),
         Edge("Up", "Up", type = "null", K = 3),
         StartEnd(start = "Dw", end = "Dw"))
gfpop(data =  myData, mygraph = myGraph, type = "mean")
```

If we skip all these constraints and use a standard fpop algorithm, the result is the following

```{r}
myGraphStd <- graph(penalty = 2*log(n), type = "std")
gfpop(data =  myData, mygraph = myGraphStd, type = "mean")
```

### abs edge

With a unique `"abs"` edge, we impose a difference between the means of size at least 1.  


```{r}
n <- 10000
myData <- dataGenerator(n, c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1), c(0, 1, 0, 2, 1, 2, 0, 1, 0, 1), sigma = 0.5)
beta <- 2*log(n)
myGraph <- graph(
  Edge(0, 0,"abs", penalty = beta, gap = 1),
  Edge(0, 0,"null"))
gfpop(data =  myData, mygraph = myGraph, type = "mean")
```

Notice that some of the edges are forced, the vector `forced` contains non-zero values.


### Exponential decay

The null edge corresponds to an exponential decay state if its parameter is not equal to 1. 

```{r}
n <- 1000
mydata <- dataGenerator(n, c(0.2, 0.5, 0.8, 1), c(5, 10, 15, 20), sigma = 1, gamma = 0.966)
beta <- 2*log(n)
myGraphDecay <- graph(
  Edge(0, 0, "up", penalty = beta),
  Edge(0, 0, "null", 0, decay = 0.966)
  )
g <- gfpop(data =  mydata, mygraph = myGraphDecay, type = "mean")
g
```


and we plot the result 

```{r, expdecay}
gamma <- 0.966
len <- diff(c(0, g$changepoints))
signal <- NULL
for(i in length(len):1)
  {signal <- c(signal, g$parameters[i]*c(1, cumprod(rep(1/gamma,len[i]-1))))}
signal <- rev(signal)

ylimits <- c(min(mydata), max(mydata))
plot(mydata, type ='p', pch ='+', ylim = ylimits)
par(new = TRUE)
plot(signal, type ='l', col = 4, ylim = ylimits, lwd = 3)
```


<a id="gc"></a>

## Graph construction

In the `gfpop` package, graphs are represented by a dataframe with 9 features and build with the R functions `Edge`, `Node`, `StartEnd` and `graph`.


```{r}
emptyGraph <- graph()
emptyGraph
```


`state1` is the starting node of an edge, `state2` its ending node. `type` is one of the available edge type (`"null"`, `"std"`, `"up"`, `"down"`, `"abs"`). `penalty` is a nonnegative parameter: the additional cost $\beta_i$ to consider when we move within the graph using a edge (or stay on the same node). `parameter` is annother nonnegative parameter, a characteristics of the edge, depending of its type (it is a decay if type is "null" and a gap otherwise). `K` and `a` are robust parameters. `min` and `max` are used to constrain the rang of value for the node parameter.

We add edges into a graph as follows

```{r}
myGraph <- graph(
  Edge("E1", "E1", "null"),
  Edge("E1", "E2", "down", 3.1415, gap = 1.5)
)
myGraph
```

we can only add edges to this dataframe using the object `Edge`.

The graph can contain information on the starting and/or ending edge to use with the `StartEnd` function. 

```{r}
beta <- 2 * log(1000)
myGraph <- graph(
  Edge("Dw", "Dw", "null"),
  Edge("Up", "Up", "null"),
  Edge("Dw", "Up", "up", penalty = beta, gap = 1),
  Edge("Dw", "Dw", "down", penalty = beta),
  Edge("Up", "Dw", "down", penalty = beta),
  StartEnd(start = "Dw", end = "Dw"))
myGraph
```


Some graphs are often used: they are defined by default in the `graph` function. To use these graphs, we specify a string `type` equal to `"std"`, `"isotonic"`, `"updown"` or `"relevant"`.
For example,



```{r}
myGraphIso <- graph(penalty = 12, type = "isotonic")
myGraphIso
```

The function `Node` can be used to restrict the range of value for parameter associated to a node (called also a vertex). For example the following graph is an isotonic graph with inferred parameters between 0 et 1 only.


```{r}
myGraph <- graph(
  Edge("Up", "Up", "up", penalty = 3.1415),
  Edge("Up", "Up"),
  Node("Up", min = 0, max = 1)
  )
myGraph
```

<a id="suppl"></a>

## Supplementary R functions

### Data generator function

the `dataGenerator` function is used to simulate `n` data-points from a distribution of `type` equal to `"mean"`, `"poisson"`, `"exp"`, `"variance"` or `"negbin"`. Standard deviation parameter `sigma` and decay `gamma` are specific to the Gaussian mean model. `size` is linked to the R `rnbinom` function from R stats package.

### Standard deviation estimation

We often need to estimate the standard deviation from the observed data to normalize the data or choose the edge penalties. The `sdDiff` returns such an estimation with the default HALL method [Hall et al., 1990] well suited for time series with change-points.


[Back to Top](#top)

